Do More Subdivisions in Wards Increase Voter Turnout? An Analysis of the 2022 Toronto Municipal Election

### 2023-06-03 outputs/paper/paper.qmd
Added more content to the discussion section, including measurement errors and limitations

The Wards within Scarborough (Wards 20-25) consistently had some of the lowest voter turnouts and fewest number of subdivisions across all 25 wards. Ward 23 (Scarborough North) witnessed the third lowest voter turnout across the entire city at 25% and the fewest number of subdivisions at 38. Ward 20 (Scarborough Southwest) had the highest turnout of all Scarborough wards at 32% and had 61 subdivisions. Scarborough's challenges with accessible public transit are well recognized, potentially contributing to the lower voter turnout.

The creation of subdivision boundaries within wards lacks transparency and clarity. The City Clerk is responsible for drawing up all ward subdivisions. We don’t know what data and information they use to make these decisions and what types of bias may be introduced. Don’t know if there’s a public input process or if like creating ward boundaries, it is an individual who has little local knowledge of the very community impacted by the creation of boundaries

Toronto Elections [@ElxnAccessibility] clearly outlines their process for finding and establishing accessible polling locations within subdivisions but does not elaborate on the process of establishing subdivisions themselves.

There is a lack of clarity whether the people who voted at Advanced Polls or by Mail In Ballot were removed from the total number of electors by subdivision, since these individuals did not vote at their assigned subdivision. This could have an impact on the total number of voters on a subdivision and ward basis.

There are several limitations with conducting this analysis. Number of subdivisions and polling location placement alone cannot solely dictate voter behaviour, as there are many other socio-economic factors which inform why an elector does or does not cast a ballot.

## Conclusion
This paper investigates the number of subdivisions within Toronto’s 25 wards in relation to voter turnout to see if the wards with more subdivisions had higher turnout during the 2022 municipal election. My results reveal that wards with more subdivisions did not correlate to higher voter turnout. Factors such as income, level of education attained, and an analysis of local races for councillor and school board trustee within a ward (whether there was an incumbent or not) should be analyzed in further studies of voter behaviour during municipal elections. Data from the upcoming Toronto Mayoral by-election on June 26, 2023 may reveal further insights into the placement of polling locations in relation to voter turnout. Cities and municipalities across Ontario are facing declining voter turnout and the factors must be identified and addressed, whether it be more strategically choosing subdivisions and polling locations in wards, introducing online voting mechanisms, or implementing democratic reform [@reducingcost]. Strong local representation is crucial for the future of democracy in Toronto and subdivisions play an important role in ensuring voters have their voices heard.

>>>>

The Wards within Scarborough (Wards 20-25) consistently had some of the lowest voter turnouts and fewest number of subdivisions across all 25 wards. Ward 23 (Scarborough North) witnessed the third lowest voter turnout across the entire city at 25% and the fewest number of subdivisions at 38. Ward 20 (Scarborough Southwest) had the highest turnout of all Scarborough wards at 32% and had 61 subdivisions. Scarborough's recognized challenges with accessible public transit, paired with fewer subdivisions and polling locations potentially contribute to the lower voter turnout [@transitdesert].

Within the data set used for analysis, there are several potential measurement errors. Firstly, the raw data set listed the total number of subdivisions per ward as "NA", despite other aspects of voter turnout and subdivisions being calculated by Toronto Elections. Secondly, there were gaps in the number and corresponding numbering of subdivisions which were not explained by the City Clerk or Toronto Elections. For example, Ward 1 is listed as having 60 subdivisions, plus Advanced Vote and Mail In Voting locations, but subdivisions 4, 20, 43-46, 50, and 58 are unaccounted for, totally 51 subdivisions. Thirdly, there is a lack of clarity whether the people who voted at Advanced Polls or by Mail In Ballot were removed from the total number of electors by subdivision, since these individuals did not vote at their assigned subdivision. If voters cast their ballot early and were not removed from the total number of electors at their specific subdivision, this could impact how voter turnout by subdivision is calculated.

There are several limitations with conducting this analysis. The creation of subdivision boundaries within wards lacks transparency and clarity, while impacting the local democratic process. The City Clerk is responsible for drawing up all ward subdivisions, however the data and information they use to make these decisions is not public and it is challenging to account for any bias introduced during this process. Toronto Elections @ElxnAccessibility clearly outlines their process for finding and establishing accessible polling locations within subdivisions but does not elaborate on the process of establishing subdivisions themselves.

Moreover, the number of subdivisions and polling location placement alone cannot dictate voter behaviour, as there are many other socio-economic factors which inform why an elector does or does not cast a ballot. The caliber of local candidates and the nature of the election, whether an incumbent is standing for re-election or the seat is open for a new councillor impacts voter turnout should be accounted for in future analysis.

## Conclusion
This paper investigates the number of subdivisions within Toronto’s 25 wards in relation to voter turnout to see if the wards with more subdivisions had higher turnout during the 2022 municipal election. My results reveal that wards with more subdivisions did not correlate to having higher voter turnout. Factors such as income, level of education attained, and an analysis of local races for councillor and school board trustee within a ward (whether there was an incumbent or not) should be analyzed in further studies of voter behaviour during municipal elections. Data from the upcoming Toronto Mayoral by-election on June 26, 2023 may reveal further insights into the placement of polling locations in relation to voter turnout. Cities and municipalities across Ontario are facing declining voter turnout and the factors must be identified and addressed, whether it be more strategically choosing subdivisions and polling locations in wards, introducing online voting mechanisms, or implementing democratic reform [@reducingcost]. Strong local representation is crucial for the future of democracy in Toronto and subdivisions play an important role in ensuring voters have their voices heard.
